---
title: "PRELIMINARY ANALYSIS"
date: "2025-04-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
# load the required packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
library(tidyr) 
library(tibble)
library(readxl)
library(kableExtra)
library(ggbeeswarm)
library(RColorBrewer)
library(readstata13)
## Set a working directory
setwd("/Manuscript_Analysis_03_2025")
```

### Post Moire Analysis

#### Organise and merge data from the 3 phases. Do QC on them before running MOIRE and Dcifer (Muko and Maziba)
#### Here, I will consider MOI, Effective MOI, Heterozygosity, and Within-host relatedness
```{r, warning=FALSE, message=FALSE, echo=FALSE}
## Reading in these csv files for each site
### How many samples per Round: Maziba and Muko
 
### Round 1
Muko_R1<- read.csv("Data/raw/Muko_R1.csv")
Maziba_R1<- read.csv("Data/raw/Maziba_R1.csv")

## Samples
length(unique(Muko_R1$SampleID)) # 86 samples
length(unique(Maziba_R1$SampleID)) # 25 samples

### Round 2
Muko_R2<- read.csv("Data/raw/Muko_R2.csv")
Maziba_R2<- read.csv("Data/raw/Maziba_R2.csv")

## Samples
length(unique(Muko_R2$SampleID)) # 34 samples
length(unique(Maziba_R2$SampleID)) # 27  samples

### Round 3
Muko_R3<- read.csv("Data/raw/Muko_R3.csv")
Maziba_R3<- read.csv("Data/raw/Maziba_R3.csv")

## Samples
length(unique(Muko_R3$SampleID)) # 52 samples
length(unique(Maziba_R3$SampleID)) # 43  samples

########### Bind together


Maziba_all <- bind_rows(Maziba_R1, Maziba_R2, Maziba_R3)

length(unique(Muko_all$SampleID)) # 172
length(unique(Maziba_all$SampleID))# 95

#add data across samples, drop "allele" because it is different across seq runs
# Muko

Muko_data <- Muko_all %>% 
  group_by(SampleID, Locus, ASV, PseudoCIGAR) %>% 
  summarise(Reads = sum(Reads), .groups = "drop") %>% 
  ungroup()

# Maziba
Maziba_data <- Maziba_all %>% 
 group_by(SampleID, Locus, ASV, PseudoCIGAR) %>% 
  summarise(Reads = sum(Reads), .groups = "drop") %>% 
  ungroup()

# Check for this sample "IM-23-120-BIIV"
print(ifelse("IM-23-120-BIIV" %in% Muko_data$SampleID, "Sample is present.", "Sample is NOT present."))
```

#### QC the per site merged data 

##### Muko Samples
```{r,}
## Muko
### Select only data from Pool A
Muko_data <- Muko_data %>%
  filter(endsWith(Locus, "A"))
length(unique(Muko_data$SampleID))

### Filter out loci that with 10 reads and below
Muko_data <- Muko_data %>% 
  filter(Reads > 10)
## Exclude these three loci, highly associated with false positive
excluded_loci <- c("Pf3D7_05_v3-550803-551072-1A", "Pf3D7_07_v3-1028330-1028603-1A", "Pf3D7_11_v3-1374917-1375148-1A") 
filtered_data<- Muko_data %>%
  filter(!Locus %in% excluded_loci) 

length(unique(filtered_data$SampleID)) ## 172 samples

## Filter out samples based % of loci with 100 read.

# Group by SampleID and count the number of loci with reads >= 100 for each sample
sample_counts <- filtered_data%>%
  group_by(SampleID) %>%
  summarise(perc_loci_above_100 = sum(Reads >= 100) / n()) %>%
  ungroup()

## We shall consider 50% of loci having at least 100 reads
# Retaining samples: 50% loci having reads >= 100
filtered_samples_50 <- sample_counts %>%
  filter(perc_loci_above_100 >= 0.50) ## This retains 164 samples
#length(unique(filtered_samples_50$SampleID))

# Extract sample IDs that do not meet the criteria
rejected_sample_ids <- sample_counts %>%
  filter(perc_loci_above_100 < 0.5) %>%  
  select(SampleID)

### Filter to remove those that did not pass the criteria of 50% loci having 100 reads
filtered_data<- filtered_data %>%
  filter(!SampleID %in% rejected_sample_ids$SampleID) 
length(unique(filtered_data$SampleID))

#Remember pseudocigars "mask" certain parts of the ASV that are unreliable 

filtered_muko <- filtered_data %>% 
 # select(-ASV) %>% 
  group_by(SampleID, Locus, PseudoCIGAR) %>% #drops out allele/ASV, which is OK 
  dplyr::summarize(Reads = sum(Reads)) %>% #this combines reads in the Locus that have matching Pseudocigars 
  ungroup () %>% 
   group_by(SampleID, Locus) %>% #now calculate the norm.reads.allele by PseudoCIGAR 
    mutate(norm.reads.allele = Reads/sum(Reads)) %>% 
    filter(norm.reads.allele > 0.01) %>% #filter out 
  mutate(
    sum_reads = sum(Reads), 
    norm.reads.allele = Reads/ sum_reads)

length(unique(filtered_muko$SampleID)) # 164 samples

## Cross-checking the calculated allele frequencies
check_freq <- filtered_muko %>%
  group_by(SampleID, Locus) %>%
  summarise(freq_sum = sum(norm.reads.allele), .groups = "drop") %>%
  mutate(is_ok = abs(freq_sum - 1) < 0.001) # Allowing a small rounding error
##-----------------------------------------------------------

# Rename some columns to have moire format
filtered_allele_data <- dplyr::rename(filtered_muko, sample_id = SampleID)
filtered_allele_data <- filtered_muko  %>% dplyr::rename(sample_id = SampleID)
filtered_allele_data <- filtered_allele_data  %>% dplyr::rename(locus = Locus)
filtered_allele_data <- filtered_allele_data  %>% dplyr::rename(allele = PseudoCIGAR)

## Filter out sample "IM-23-120-BIIV" since its low qc values
print(ifelse("IM-23-120-BIIV" %in% filtered_allele_data$sample_id, "Sample is present.", "Sample is NOT present."))

# "Sample is NOT present." The sample was excluded by the filtering strategies above. But we can explicitly exclude it again
filtered_allele_data <- filtered_allele_data[filtered_allele_data$sample_id != "IM-23-120-BIIV", ]
length(unique(filtered_allele_data$sample_id))

### Let's remove "Duplicated samples"
Duplic_samples<- read.csv("Data/raw/duplicated_samples_to_remove_from_analysis.csv") %>% dplyr::rename(sample_id=SampleID)
##
filtered_allele_data_mu <- filtered_allele_data %>%
  anti_join(Duplic_samples, by = "sample_id")
length(unique(filtered_allele_data_mu$sample_id))

##----------
write.csv(filtered_allele_data_mu, "Data/processed/Muko_filtered_data.csv", row.names = FALSE)
```

#### QC on Maziba samples

```{r,}
### Retain only data from Pool A
Maziba_data <- Maziba_data %>%
  filter(endsWith(Locus, "A")) ##  
length(unique(Maziba_data$SampleID))

### Filter out loci that with 10 reads and below
Maziba_data <- Maziba_data %>% 
  filter(Reads > 10)
## Exclude these three loci, highly associated with false positive
excluded_loci <- c("Pf3D7_05_v3-550803-551072-1A", "Pf3D7_07_v3-1028330-1028603-1A", "Pf3D7_11_v3-1374917-1375148-1A") 
filtered_data<- Maziba_data %>%
  filter(!Locus %in% excluded_loci) 

length(unique(filtered_data$SampleID)) ## 95 samples

## Filter out samples based % of loci with 100 read.

# Group by SampleID and count the number of loci with reads >= 100 for each sample
sample_counts <- filtered_data%>%
  group_by(SampleID) %>%
  summarise(perc_loci_above_100 = sum(Reads >= 100) / n()) %>%
  ungroup()

## We shall consider 50% of loci having at least 100 reads
# Retaining samples: 50% loci having reads >= 100
filtered_samples_50 <- sample_counts %>%
  filter(perc_loci_above_100 >= 0.50) ## This retains 94 samples
length(unique(filtered_samples_50$SampleID))

# Extract sample IDs that do not meet the criteria
rejected_sample_ids <- sample_counts %>%
  filter(perc_loci_above_100 < 0.5) %>%  
  select(SampleID)

### Filter to remove those that did not pass the criteria of 50% loci having 100 reads
filtered_data<- filtered_data %>%
  filter(!SampleID %in% rejected_sample_ids$SampleID) 
length(unique(filtered_data$SampleID))

### Filter by allele frequency: loci with allele frequencies below 1% to be excluded.
## Using pseudoCIGAR as the true allele since it masks certain parts of ASVs that are unreliable
filtered_maziba <- filtered_data %>% 
 group_by(SampleID, Locus, PseudoCIGAR) %>% #drops out allele/ASV, which is OK 
  dplyr::summarize(Reads = sum(Reads)) %>% #this combines reads in the Locus that have matching Pseudocigars 
  ungroup () %>% 
   group_by(SampleID, Locus) %>% #now calculate the norm.reads.allele by PseudoCIGAR 
    mutate(norm.reads.allele = Reads/sum(Reads)) %>% 
    filter(norm.reads.allele > 0.01) %>% #filter out 
  mutate(
    sum_reads = sum(Reads), 
    norm.reads.allele = Reads/ sum_reads)
#
length(unique(filtered_maziba$SampleID)) # 94 samples
#
## Cross-checking the calculated allele frequencies
check_freq <- filtered_maziba %>%
  group_by(SampleID, Locus) %>%
  summarise(freq_sum = sum(norm.reads.allele), .groups = "drop") %>%
  mutate(is_ok = abs(freq_sum - 1) < 0.001) # Allowing a small rounding error

# Rename some columns to have moire format
filtered_allele_data <- filtered_maziba  %>% dplyr::rename(sample_id = SampleID)
filtered_allele_data <- filtered_allele_data  %>% dplyr::rename(locus = Locus)
filtered_allele_data <- filtered_allele_data  %>% dplyr::rename(allele = PseudoCIGAR)

#
print(ifelse("IM-23-120-BIIV" %in% filtered_allele_data$sample_id, "Sample is present.", "Sample is NOT present."))

# Remove duplicated samples
filtered_allele_data_ma <- filtered_allele_data %>%
  anti_join(Duplic_samples, by = "sample_id")
length(unique(filtered_allele_data_ma$sample_id))
## 
write.csv(filtered_allele_data_ma, "Data/processed/Maziba_filtered_data.csv", row.names = FALSE)
```

### Kamwezi
```{r,}
## Kamwezi data
Kamwezi<- read.csv("Data/raw/Kamwezi_allele_data.csv")
##
Kamwezi <- Kamwezi %>% 
  group_by(SampleID, Locus, ASV, PseudoCIGAR) %>% 
  summarise(Reads = sum(Reads), .groups = "drop") %>% 
  ungroup()

## QC
### Select only data from Pool A
Kamwezi_data <- Kamwezi %>%
  filter(endsWith(Locus, "A"))
length(unique(Kamwezi_data$SampleID))

### Filter out loci that with 10 reads and below
Kamwezi_data <- Kamwezi_data %>% 
  filter(Reads > 10)
## Exclude these three loci, highly associated with false positive
excluded_loci <- c("Pf3D7_05_v3-550803-551072-1A", "Pf3D7_07_v3-1028330-1028603-1A", "Pf3D7_11_v3-1374917-1375148-1A") 
filtered_data<- Kamwezi_data %>%
  filter(!Locus %in% excluded_loci) 

length(unique(filtered_data$SampleID)) ## 100 samples

## Filter out samples based % of loci with 100 read.

# Group by SampleID and count the number of loci with reads >= 100 for each sample
sample_counts <- filtered_data%>%
  group_by(SampleID) %>%
  summarise(perc_loci_above_100 = sum(Reads >= 100) / n()) %>%
  ungroup()

## We shall consider 50% of loci having at least 100 reads
# Retaining samples: 50% loci having reads >= 100
filtered_samples_50 <- sample_counts %>%
  filter(perc_loci_above_100 >= 0.50) ## This retains 164 samples
#length(unique(filtered_samples_50$SampleID))

# Extract sample IDs that do not meet the criteria
rejected_sample_ids <- sample_counts %>%
  filter(perc_loci_above_100 < 0.5) %>%  
  select(SampleID)

### Filter to remove those that did not pass the criteria of 50% loci having 100 reads
filtered_data<- filtered_data %>%
  filter(!SampleID %in% rejected_sample_ids$SampleID) 
length(unique(filtered_data$SampleID))

#Remember pseudocigars "mask" certain parts of the ASV that are unreliable 

filtered_kamwezi <- filtered_data %>% 
 group_by(SampleID, Locus, PseudoCIGAR) %>% #drops out allele/ASV, which is OK 
  dplyr::summarize(Reads = sum(Reads)) %>% #this combines reads in the Locus that have matching Pseudocigars 
  ungroup () %>% 
   group_by(SampleID, Locus) %>% #now calculate the norm.reads.allele by PseudoCIGAR 
    mutate(norm.reads.allele = Reads/sum(Reads)) %>% 
    filter(norm.reads.allele > 0.01) %>% #filter out 
  mutate(
    sum_reads = sum(Reads), 
    norm.reads.allele = Reads/ sum_reads)
#
length(unique(filtered_kamwezi$SampleID)) # 100 samples

# Rename some columns to have moire format
filtered_allele_data <- filtered_kamwezi  %>% dplyr::rename(sample_id = SampleID)
filtered_allele_data <- filtered_allele_data  %>% dplyr::rename(locus = Locus)
filtered_allele_data <- filtered_allele_data  %>% dplyr::rename(allele = PseudoCIGAR)

length(unique(filtered_allele_data$sample_id))
### Let's remove "Duplicated samples"
Duplic_samples<- read.csv("Data/raw/duplicated_samples_to_remove_from_analysis.csv") %>% dplyr::rename(sample_id=SampleID)
##
filtered_allele_data_ka <- filtered_allele_data %>%
  anti_join(Duplic_samples, by = "sample_id")
length(unique(filtered_allele_data$sample_id))

##----------
write.csv(filtered_allele_data_ka, "Data/processed/Kamwezi_filtered_data.csv", row.names = FALSE)
```

### Retaining data that was collected up to June, 2024 for Muko and Maziba

```{r,}
Muko_IDs <- filtered_allele_data_mu[, "sample_id", drop = FALSE] %>% unique()
Maziba_IDs <- filtered_allele_data_ma[, "sample_id", drop = FALSE] %>% unique()
#
LTS <- bind_rows(Muko_IDs, Maziba_IDs)
length(unique(LTS$sample_id))
## Let's extract dates from the survey and umsp database
Survey <- read.csv("Data/raw/Muko_Maziba_travel_survey_data.csv") %>% dplyr::rename(sample_id=barcode) %>% select(sample_id, date)

#
LTS_dates <- LTS %>%
  left_join(Survey, by = "sample_id")

sum(is.na(LTS_dates$date))

## Umsp database
UMSP_database <- read.dta13("Data/raw/2025_01_26_UMSP_low_trans_sites_and_Kamwezi_all_data.dta",
           convert.factors = TRUE, generate.factors = TRUE,
           encoding = "UTF-8", fromEncoding = NULL, 
           convert.underscore = FALSE, missing.type = TRUE,
           convert.dates = TRUE, replace.strl = TRUE,
           add.rownames = FALSE, nonint.factors = TRUE,
           select.rows = NULL, select.cols = NULL, strlexport =
             FALSE, strlpath = ".") %>%
  dplyr::rename(sample_id=barcode) 

## UMSP data
UMSP <- UMSP_database %>%
  filter(site %in% c("Muko", "Maziba") & format(date, "%Y") == "2023") %>% select(sample_id, date) %>% droplevels()

#
LTS_dates <- LTS_dates %>%
  left_join(UMSP %>% select(sample_id, date), by = "sample_id", suffix = c("", "_umsp"))%>%
  mutate(date = coalesce(as.Date(date), as.Date(date_umsp))
  ) %>%  select(-date_umsp)

sum(is.na(LTS_dates$date)) # 57 missing dates

# Let's arrange date column
LTS_ordered <- LTS_dates %>%
  arrange(date)

#
LTS_filtered <- LTS_ordered %>%
  filter(as.Date(date) <= as.Date("2024-06-30"))
#-------------------------------------------------------------------------------------
### Save the filtered data per site
# Muko
filtered_allele_data_mu <- filtered_allele_data_mu %>%
  semi_join(LTS_filtered, by = "sample_id")

length(unique(filtered_allele_data_mu$sample_id))
#
write.csv(filtered_allele_data_mu, "Data/processed/Muko_filtered_data1.csv", row.names = FALSE)
##________________________________________________________________________________________________

## Maziba
filtered_allele_data_ma <- filtered_allele_data_ma %>%
  semi_join(LTS_filtered, by = "sample_id")

length(unique(filtered_allele_data_ma$sample_id))
#
write.csv(filtered_allele_data_ma, "Data/processed/Maziba_filtered_data1.csv", row.names = FALSE)
```

### Lets drop duplicated samples from the Survey data
```{r,}

Survey <- read.csv("Data/raw/Muko_Maziba_travel_survey_data.csv") %>% dplyr::rename(sample_id=barcode)

faked_samples<- read.csv("Data/raw/samples_to_keep_in_geno_analysis_but_throw_out_travel_data (1).csv") %>% dplyr::rename(sample_id=SampleID)

## Filter out this samples from the survey
Survey_filtered <- Survey %>%
  anti_join(faked_samples, by = "sample_id")
## 
Survey_filtered <- Survey_filtered %>%
  filter(as.Date(date) <= as.Date("2024-06-30"))

write.csv(Survey_filtered, "Data/processed/Survey_filtered_data.csv", row.names = FALSE)
```



