---
title: "Dcifer Analysis: Figure 3"
author: "Monica Mbabazi"
date: "2025-08-19th"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls(all=TRUE))
## Loading required libraries
library(dcifer)
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(readr)
library(plotly)
library(ggnewscale)
library(tidyverse)
library(descriptr)
library(ggraph)
library(igraph)
library(grid)
library(kableExtra)
library(cluster)  # For kmeans clustering
library(mclust) 
library(tidygraph)
library(gridExtra)
library(cowplot)
library(png)
library(grid)
library(forcats)
library(ggplot2)
library(scales)
library(ggnewscale)
setwd("/home/monica/Desktop/IMMRSE_Study/IMMRSE-Concept/Manuscript_Analysis_03_2025/")
```

#### Running Dcifer on Muko Data (Fig3)
```{r, echo=TRUE}
## Loading  and reading the data for Dcifer
## Moire outputs
muko_moire_results <- readr::read_rds("MOIRE_RESULTS/Muko/Muko_mcmc_results.rds")

## Moire data
muko_data_moire <- read.csv("Filtered_data/Muko_filtered_TNM_data.csv", header = T)
#muko_data_moire <- muko_data_moire %>% rename("allele"="PseudoCIGAR")
# Format the data
muko_data <- moire::load_long_form_data(muko_data_moire)

# Function to calculate median allele frequencies
calculate_median_afs <- function(x) {
  median_afs <- lapply(x$chains[[1]]$allele_freqs, function(af) {
    mat <- matrix(unlist(af), ncol = length(af))
    d <- dist(t(mat))
    af[[which.min(rowSums(as.matrix(d)))]]
  })
  return(median_afs)
}
# Calculate median allele frequencies for each site
muko_median_afs <- calculate_median_afs(muko_moire_results)

# Define uninformative loci for each site
mu_uninformative_loci <- muko_data$uninformative_loci

# Filter data for each site
filtered_muko <- muko_data_moire %>% filter(!locus %in% mu_uninformative_loci)

# Prepare DCIFER input for each site
muko_dsmp<- dcifer::formatDat(filtered_muko, "sample_id", "locus", "allele")
### COI
muko_coi   <- moire::summarize_coi(muko_moire_results)$post_coi_med

### Median frequency
muko_dsmp <- dcifer::matchAfreq(muko_dsmp, muko_median_afs)

## IBD per site
dres_df <-ibdDat(muko_dsmp, muko_coi, muko_median_afs, pval = TRUE, confint = TRUE, rnull = 0.125, alpha = 0.05, nr = 1e3)
```

#### Highly related pairs
```{r,}
# Extract IBD estimates from rnull = 0.25 run
ibd_est <- dres_df[, , "estimate"]
ibd_lower <- dres_df[, , "CI_lower"]
ibd_upper <- dres_df[, , "CI_upper"]

#rnull = 0.125
ibd_df <- as.data.frame(ibd_est)
ibd_df <- tibble::rownames_to_column(ibd_df, var = "x")

ibd_lower_df <- as.data.frame(ibd_lower)
ibd_lower_df <- tibble::rownames_to_column(ibd_lower_df, var = "x")

ibd_upper_df <- as.data.frame(ibd_upper)
ibd_upper_df <- tibble::rownames_to_column(ibd_upper_df, var = "x")

#this is a DF with all ibd data in longform between non-matching pairs of samples
ibd_df_long <- ibd_df %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "ibd"
  ) %>% filter(!is.na(ibd)) 

hist(ibd_df_long$ibd)

#### BIND IN CI VALUES TO EACH DF ###############

CI_lower_df_long <- ibd_lower_df %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "ibd_lower")

CI_upper_df_long <- ibd_upper_df %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "ibd_upper")

#bind CIs into long df 
ibd_df_long <- ibd_df_long %>% 
  left_join(CI_lower_df_long) %>% 
  left_join(CI_upper_df_long)

#### BIND IN P VALUES TO EACH DF ###############
pvals <- dres_df[, , "p_value"]
pvals <- as.data.frame(pvals)
pvals <- tibble::rownames_to_column(pvals, var = "x")

#pval 0.05
pvals_df_long <- pvals %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "p_value"
  ) %>% filter(!is.na(p_value)) %>% 
  mutate(p_value = p_value/2) #divide by 2 for 1 sided test

#bind pvals into long df 
ibd_df_long <- ibd_df_long %>% 
  left_join(pvals_df_long)


#make p-value very large if r <0.125

ibd_df_long <- ibd_df_long %>% mutate(p_value = ifelse(ibd <0.125, 1, p_value))

#here is where i do corrections for rnull = 0.25 dataframe
long_ibd1 <- ibd_df_long %>% 
  mutate(p_BH = p.adjust(p_value, method = "BH"), 
         p_BF = p.adjust(p_value, method = "bonferroni")) %>% 
  mutate(sig_0.01 = ifelse(p_value < 0.01, 1, 0), sig_0.01_BH = ifelse(p_BH < 0.01, 1, 0), 
         sig_0.01_BF = ifelse(p_BF < 0.01, 1, 0)) %>% 
  mutate(sig_0.05 = ifelse(p_value < 0.05, 1, 0), sig_0.05_BH = ifelse(p_BH < 0.05, 1, 0),
         sig_0.05_BF= ifelse(p_BF < 0.05, 1, 0)) %>% 
  mutate(n = 1)

#highly related pairs? 
highly_related_df <- long_ibd1 %>% filter(sig_0.05_BH == 1)

write.csv(highly_related_df, 
          file = "Final_Manuscript_Figures/Re-analysis_Figs_2026/Muko_highly_rel-pairs.csv", 
          row.names = FALSE)
#highly related pairs out of all pairwise comparisons # 0.001724138 (0.17%)
sum(long_ibd1$sig_0.05_BH)/sum(long_ibd1$n)

range(highly_related_df$ibd, na.rm = TRUE) # 0.562 - 1.000
```

### Network plots based on DCIFER

```{r,}
# Extract IBD estimates for each site
dmat_muko <- dres_df[, , "estimate"]
all_samples <- rownames(dmat_muko)
g <- graph_from_data_frame(highly_related_df, vertices = data.frame(name = all_samples), directed = FALSE)

##
clusters <- components(g)
## Identify which samples are in a cluster with more than one node
cluster_sizes <- table(clusters$membership)
is_singleton_cluster <- cluster_sizes == 1

## Re-assign singletons to a shared group (e.g., "Singleton")
cluster_labels <- clusters$membership
cluster_labels[is_singleton_cluster[cluster_labels]] <- "Singleton"

tg <- as_tbl_graph(g) %>%
  mutate(cluster = as.factor(cluster_labels))

# Plot
ggraph(tg, layout = "fr") +
  geom_edge_link(alpha = 0.3) +
  geom_node_point(aes(color = cluster), size = 2) +
  labs(title = "Clusters of Highly Related Samples ",
       color = "Cluster ID") +
  theme_minimal()

## Plot 2
library(scales)

## Build custom colors
unique_clusters <- sort(unique(cluster_labels[cluster_labels != "Singleton"]))
cluster_colors <- setNames(
  c("gray", rainbow(length(unique_clusters))),
  c("Singleton", unique_clusters)
)

# Plot
p2 <-ggraph(tg, layout = "fr") +
  geom_edge_link(alpha = 0.3) +
  geom_node_point(aes(color = cluster), size = 3) +
  scale_color_manual(values = cluster_colors) +
  labs(title = "Clusters of Highly Related Samples",
       color = "Cluster ID") +
  theme_minimal()+
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA))
p2
#ggsave("DCIFER_RESULTS/Muko_Dcifer_Network_plot.png", p2, width = 10, height = 8, units = "in")

```

## Transmission Network Plots

```{r, }
### RESISTANCE DATA FOR Muko
Res_genotypes <- read_csv("Resistance_Analysis/Resistance_Data/sample_genotypes_all.csv") #%>% select(-c(k13_C469Y, k13_C469F))
# Keep only samples from Maziba
Res_genotypes <- Res_genotypes %>% 
  filter(Sample.Site == "Muko") %>% dplyr::rename(sample_id=barcode)

##### TABULATE Ns #####
meta_long <- pivot_longer(Res_genotypes, cols = arps10_V127M:k13_C469F, names_to = "SNP", values_to = "Genotype") 

#k13 only
meta_long <- meta_long %>% filter(grepl("k13", SNP))
###----------------------------------------------------------------------------------
k13_P441L <- meta_long %>% filter(SNP == "k13_P441L") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(P441L = Genotype)

k13_675 <- meta_long %>% filter(SNP == "k13_A675V") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_675 = Genotype)

k13_C469Y <- meta_long %>% filter(SNP == "k13_C469Y") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_C469Y = Genotype)

k13_C469F <- meta_long %>% filter(SNP == "k13_C469F") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_C469F = Genotype)
##--------------------------------------------------------------------------------
#### Travel data 
travel <- read_csv("Data/Survey_filtered_data.csv", 
                   col_types = cols(starttime = col_date(format = "%Y-%m-%d"), 
                                    date = col_date(format = "%Y-%m-%d"))) %>%
  filter(mrccode == "Muko") %>%
  dplyr::select(sample_id, mrccode, dateofnight1st1, dateofnightlast1, date, ageyrs, agemonths, gender, villageresidence, parishresidence, districtresidence, 
         furthesttravel, travelreason1, villagedestination1, parishdestination1, districtdestination1) %>% 
  mutate(ageyrs = ifelse(is.na(ageyrs), agemonths/12, ageyrs)) %>% 
  dplyr::select(-agemonths) %>% 
  dplyr::rename(id = sample_id)

table(travel$furthesttravel, useNA = "always") 
table(travel$furthesttravel, travel$mrccode, useNA = "always") 

travel1 <- travel %>% 
  mutate(furthesttravel = ifelse(furthesttravel == "Inside village", "No travel", furthesttravel )) %>% 
  mutate(furthesttravel = ifelse((furthesttravel == "Outside parish but inside subcounty" | furthesttravel =="Outside subcounty inside district" | furthesttravel == "Outside village but inside parish"), "Inside district", furthesttravel))

table(travel1$furthesttravel, travel1$mrccode, useNA = "always") 
#-------------------------------------------------------------------------------

travel_for_merge <- travel1 %>% dplyr::rename(sample_id = id)

##
muko_travel <- travel_for_merge %>%
  mutate(furthesttravel = ifelse(is.na(furthesttravel), "Missing data", furthesttravel)) %>%
  mutate(furthesttravel = factor(furthesttravel, levels = c(
    "Outside district inside country", 
    "Inside district", 
    "No travel", 
    "Missing data"
  ), ordered = TRUE))


table(muko_travel$furthesttravel)
levels(muko_travel$furthesttravel)

is.ordered(muko_travel$furthesttravel)
str(muko_travel$furthesttravel)

## Transmission Network data
Muko_cs <- readRDS("Transmission_Network_Analysis/muko_inferred_network_edges_cs.rds")
#
edge_output <- Muko_cs %>%
  dplyr::rename(from = parent_node, to = sample_id, prob = mean_posterior_prob)

# Remove edges from nodes with `_copy` as from — we don't want `_copy` nodes in the plot
edge_output_clean <- edge_output %>%
  filter(!str_detect(from, "_copy$"))
#
real_node_ids <- unique(c(edge_output$to, edge_output_clean$from))
                        
founder_nodes <- setdiff(edge_output_clean$from, edge_output_clean$to)
## Observed founders (no incoming edge)
observed_founders <- setdiff(edge_output_clean$from, edge_output_clean$to)

## Singleton founders (only infected by _copy and never infected by anyone else)
singleton_founders <- edge_output %>%
  filter(str_detect(from, "_copy$")) %>%
  pull(to) %>%
  setdiff(edge_output_clean$to)

# Combine both groups
all_founders <- union(observed_founders, singleton_founders)
#
graph_nodes_annotated <- tibble(name = real_node_ids) %>%
  left_join(muko_travel, by = c("name" = "sample_id")) %>%
  left_join(k13_P441L,         by = c("name" = "id")) %>%
  left_join(k13_675,       by = c("name" = "id")) %>%
  left_join(k13_C469Y,     by = c("name" = "id")) %>%
  left_join(k13_C469F,     by = c("name" = "id")) %>%
  mutate(
    founder = name %in% all_founders,
   # month_year = paste0(month(date, label = TRUE, abbr = TRUE), "_", str_sub(year(date), 3, 4)),
   month_year = paste0(lubridate::month(date, label = TRUE, abbr = TRUE), "_", stringr::str_sub(lubridate::year(date), 3, 4)),

    month_year = factor(month_year, levels = unique(month_year[order(date)]), ordered = TRUE)
  )

## Replace all NAs with "Missing data"
graph_nodes_annotated <- graph_nodes_annotated %>%
  mutate(furthesttravel = replace_na(furthesttravel, "Missing data"))

table(graph_nodes_annotated$furthesttravel)
# Create final transmission graph
final_graph <- graph_from_data_frame(
  d = edge_output_clean,
  vertices = graph_nodes_annotated,
  directed = TRUE
)
### Save this file

write.csv(graph_nodes_annotated, 
          file = "DCIFER_RESULTS/graph_nodes_annotated_muko.csv", 
          row.names = FALSE)
```

### PLotting
```{r, }
## Plot with random cluster positions
Muko_travel_plot <- ggraph(final_graph, layout = 'fr') +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Furthest Travel"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  theme_void() +
  guides(size = "none") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  ggtitle("Muko Transmission Network")

Muko_travel_plot

ggsave("DCIFER_RESULTS/Muko_Travel_plot.png", Muko_travel_plot, width = 10, height = 8, units = "in")
```

## Transmission Network Plot with fixed clusters
```{r,}
set.seed(100)
## Generate layout positions using one graph (say dcifer)
shared_layout <- create_layout(final_graph, layout = "fr")
##
shared_layout <- shared_layout %>%
  mutate(
    furthesttravel = factor(
      furthesttravel,
      levels = c(
        "Outside district inside country",
        "Inside district",
        "No travel",
        "Missing data"
      )
    )
  )

## Transmission Network Plot
Muko_travel_plot <- ggraph(shared_layout) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Travel History"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  theme_void() +
  guides(size = "none") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  ggtitle("Muko Transmission Network")

Muko_travel_plot

ggsave("DCIFER_RESULTS/Muko_TN_Clusters_fixed_travel.png",Muko_travel_plot, width = 10, height = 8, units = "in")

#### ### Extract the legend and remove it from the network plot
# Function to extract legend
get_legend_plot <- function(plot) {
  g <- ggplotGrob(plot)
  legend_index <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  g$grobs[[legend_index]]
}

# Plot WITH only two legends
Muko_plot_with_legend <- ggraph(shared_layout) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Travel History"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 6, `TRUE` = 6), guide = "none") +  # <-- no size legend
  theme_void() +
  theme(
    legend.key.size = unit(1.5, "cm"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 5)),
    shape = guide_legend(override.aes = list(size = 5))
  )

# Extract legend
legend_plot <- get_legend_plot(Muko_plot_with_legend)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

#ggsave("Final_Manuscript_Figures/Mu_Transmission_legend.png",
 #      plot = arrangeGrob(legend_plot),
  #     width = 3.5, height = 4.5, dpi = 300, bg = "white")

# Main plot WITHOUT legend
Muko_plot_clean <- Muko_plot_with_legend +
  theme(legend.position = "none")

Muko_plot_clean
ggsave("Final_Manuscript_Figures/Muko_TN_Clusters_No_Legend.png",Muko_plot_clean, width = 9, height = 9, units = "in", bg="white")
```
### Transmission networks with probability-weighted edges
```{r,}
set.seed(100)

Muko_trans_clean <- ggraph(shared_layout) +
  geom_edge_link(
    aes(alpha = prob, width = prob),
    arrow   = arrow(type = "closed", length = unit(2, "mm")),
    end_cap = circle(3, "mm"),
    colour  = "black",
    show.legend = TRUE
  ) +

  # Edge legend scales: fixed 0–1 range, same breaks for all sites
  scale_edge_alpha(
    range  = c(0.25, 1),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +
  scale_edge_width(
    range  = c(0.4, 2),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +

  # Nodes
  geom_node_point(
    aes(color = furthesttravel, shape = founder),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    name = "Travel history"
  ) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 17),
    name   = "Founder"
  ) +

  guides(
    edge_width = guide_legend(order = 1),
    edge_alpha = guide_legend(order = 1),
    color      = guide_legend(order = 2),
    shape      = guide_legend(order = 3)
  ) +
  theme_void() +
 theme(
    legend.position = "none"
  )

Muko_trans_clean

ggsave("Final_Manuscript_Figures/Muko_TN_Clusters_trans_clean.png",
       Muko_trans_clean, width = 9, height = 9, units = "in", bg = "white")
##----------------------------------------------------------------------------------------

Muko_trans_clean <- ggraph(shared_layout) +
  geom_edge_link(
    aes(alpha = prob, width = prob),
    arrow   = arrow(type = "closed", length = unit(2, "mm")),
    end_cap = circle(3, "mm"),
    colour  = "black",
    show.legend = TRUE     # <-- ENABLE LEGEND
  ) +
  
  # Edge legend scales
  scale_edge_alpha(
    range = c(0.25, 1),
    name = "Transmission\nprobability"
  ) +
  scale_edge_width(
    range = c(0.4, 2),
    name = "Transmission\nprobability"
  ) +
  
  # Node appearance
  geom_node_point(
    aes(color = furthesttravel, shape = founder),
    size = 6
  ) +
  
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    name = "Travel history"
  ) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 17),
    name   = "Founder"
  ) +
  
  # Combine the edge legends so they are not duplicated
  guides(
    edge_width = guide_legend(order = 1),
    edge_alpha = guide_legend(order = 1),
    color      = guide_legend(order = 2),
    shape      = guide_legend(order = 3)
  ) +
  
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text  = element_text(size = 11)
  )
Muko_trans_clean

ggsave("Final_Manuscript_Figures/Muko_TN_Clusters_trans_legend.png",
       Muko_trans_clean, width = 12, height = 9, units = "in", bg = "white")
##-----------------------------------------------------------------------------------
### With a standard scale for this plot
Muko_trans_clean <- ggraph(shared_layout) +
  geom_edge_link(
    aes(alpha = prob, width = prob),
    arrow   = arrow(type = "closed", length = unit(2, "mm")),
    end_cap = circle(3, "mm"),
    colour  = "black",
    show.legend = TRUE
  ) +

  # Edge legend scales: fixed 0–1 range, same breaks for all sites
  scale_edge_alpha(
    range  = c(0.25, 1),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +
  scale_edge_width(
    range  = c(0.4, 2),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +

  # Nodes
  geom_node_point(
    aes(color = furthesttravel, shape = founder),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    name = "Travel history"
  ) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 17),
    name   = "Founder"
  ) +

  guides(
    edge_width = guide_legend(order = 1),
    edge_alpha = guide_legend(order = 1),
    color      = guide_legend(order = 2),
    shape      = guide_legend(order = 3)
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title    = element_text(size = 12, face = "bold"),
    legend.text     = element_text(size = 11)
  )

Muko_trans_clean
ggsave("Final_Manuscript_Figures/Muko_TN_Clusters_trans_legend1.png",
       Muko_trans_clean, width = 12, height = 9, units = "in", bg = "white")

#### Save the legend separate
# Extract legend
legend_plot <- get_legend_plot(Muko_trans_clean)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

ggsave("Final_Manuscript_Figures/Muko_trans_clean_legend1.png",
       plot = arrangeGrob(legend_plot),
       width = 3.5, height = 5, dpi = 300, bg = "white")

```


##### Investigate futher on these clusters in Muko
```{r,}
## Muko
set.seed(100)

# 1) Create a single, fixed layout on the FULL graph
layout_full <- create_layout(final_graph, layout = "fr") %>%
  as_tibble() %>%
  select(name, x, y)

# 2) Compute degree on the FULL graph (undirected degree for “singleton” test)
deg_vec <- igraph::degree(final_graph, mode = "all")
degree_df <- tibble(name = V(final_graph)$name, degree = as.numeric(deg_vec))

# 3) Join coordinates + degree; pick the nodes to KEEP (degree > 0)
node_pos <- layout_full %>%
  left_join(degree_df, by = "name")

keep_nodes <- node_pos %>%
  filter(degree > 0) %>%
  pull(name)

# 4) Induce a subgraph of ONLY connected nodes (no singletons), but KEEP attributes
g_clusters <- induced_subgraph(final_graph, vids = V(final_graph)[name %in% keep_nodes])

# 5) Prepare manual coordinates for the subgraph (match to full layout)
coord_map <- node_pos %>%
  filter(name %in% keep_nodes) %>%
  select(name, x, y)

idx <- match(V(g_clusters)$name, coord_map$name)
manual_x <- coord_map$x[idx]
manual_y <- coord_map$y[idx]

# 6) Clean up travel label (NAs -> "Missing data") & set fixed order
#    (Do this on the vertex attributes directly to avoid surprises)
vt <- vertex_attr(g_clusters)
travel <- vt$furthesttravel
travel[is.na(travel)] <- "Missing data"
travel <- factor(
  travel,
  levels = c(
    "Outside district inside country",
    "Inside district",
    "No travel",
    "Missing data"
  ),
  ordered = TRUE
)
g_clusters <- set_vertex_attr(g_clusters, "furthesttravel", value = travel)

# (Optional) If you want to color by SITE instead, set a factor here:
# site_vals <- factor(vertex_attr(g_clusters, "site"),
#                     levels = c("Kamwezi", "Maziba", "Muko"))
# g_clusters <- set_vertex_attr(g_clusters, "site", value = site_vals)

# 7) Plot with MANUAL layout (positions are fixed), edges & alpha from 'prob'
#    Color by travel; shape by founder; hide size legend.
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Travel History"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  guides(size = "none") +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title       = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("Muko Transmission Network (clusters only)")

print(p)
##----------------------------------------------------------

#### Label founder nodes by their sample Ids
founder_labels <- tibble(
  name    = V(g_clusters)$name,
  founder = V(g_clusters)$founder,
  x       = manual_x,
  y       = manual_y
) %>%
  filter(founder == TRUE)  # keep only founders

# Plot with founder labels
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  geom_text(
    data = founder_labels,
    aes(x = x, y = y, label = name),  # use 'name' for labels
    vjust = -1,
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Travel History"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  guides(size = "none") +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title       = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("Muko Transmission Network (Clusters Only, Founders Labeled)")

print(p)

## 
ggsave("DCIFER_RESULTS/Muko_TN_Clusters_with_IDs1.png", p, width = 10, height = 8, units = "in")
##------------------------------------------------------------------------------------------

### Identify samples in each cluster

# 1. Convert to tidygraph if not already
graph_tbl <- as_tbl_graph(g_clusters)

# 2. Compute clusters (connected components)
clusters <- components(as.igraph(graph_tbl))  # returns membership info

# 3. Build node table with cluster ID
node_clusters <- tibble(
  name       = V(g_clusters)$name,
  founder    = V(g_clusters)$founder,
  furthesttravel = V(g_clusters)$furthesttravel,
  cluster_id = clusters$membership
)

# 4. Optionally, renumber clusters from 1 to N
node_clusters <- node_clusters %>%
  mutate(cluster_id = as.integer(factor(cluster_id, levels = sort(unique(cluster_id)))))
##
node_clusters <- node_clusters %>%
  arrange(cluster_id, desc(founder)) %>% rename(sample_id=name)

### Add more columns from the survey data
# Select only the columns you want from Maziba_travel
Muko_subset <- muko_travel %>%
  select(sample_id, dateofnight1st1, dateofnightlast1, date, ageyrs, gender, villageresidence, districtdestination1)

# Join
# Join to node_clusters by sample_id
node_clusters <- node_clusters %>%
  left_join(Muko_subset, by = "sample_id")

##
write.csv(node_clusters, 
          file = "DCIFER_RESULTS/Muko_node_clusters_Epi.csv", 
          row.names = FALSE)
```


## DCIFER Network Plot with fixed clusters
```{r, }
set.seed(100)
# Extract node positions from shared layout

#Extract node name and x/y
transmission_layout <- shared_layout %>% 
  dplyr::select(name, x, y)

##
tg_with_coords <- tg %>%
  activate(nodes) %>%
  left_join(transmission_layout, by = "name")

## Plot
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Transparent, thin, gray edges
  geom_edge_link(
    alpha = 0.2,
    color = "gray70",
    edge_width = 0.4
  ) +
  # Optional: Use curved edges to reduce clutter
  # geom_edge_fan(alpha = 0.2, color = "gray70", edge_width = 0.3)

  # Plot nodes by cluster
  geom_node_point(
    aes(color = cluster),
    size = 3
  ) +
  scale_color_manual(values = cluster_colors) +  # Make sure this exists
  labs(title = "Kamwezi Clusters (dcifer)", color = "Cluster ID") +
  coord_fixed() +
  # Clean white background
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

dcifer_plot

###-------------------------------------------------------
## Excluding long edges from the plot
# Extract layout (should be character node names)
transmission_layout <- shared_layout %>% dplyr::select(name, x, y)

#
tg_with_coords <- tg %>%
  activate(nodes) %>%
  left_join(transmission_layout, by = "name")

## Add travel info
nodes_tbl <- tg_with_coords %>%
  activate(nodes) %>%
  as_tibble() %>%
  left_join(muko_travel, by = c("name" = "sample_id"))

##
tg_with_coords <- tg_with_coords %>%
  activate(nodes) %>%
  left_join(muko_travel, by = c("name" = "sample_id"))
##
tg_with_coords <- tg_with_coords %>%
  activate(nodes) %>%
  mutate(
    furthesttravel = replace_na(furthesttravel, "Missing data")
  )

 # Extract edge table and node table
edges_tbl <- tg_with_coords %>%
  activate(edges) %>%
  as_tibble()

#
nodes_tbl <- tg_with_coords %>%
  activate(nodes) %>%
  as_tibble() %>%
  mutate(row_id = row_number())
#
node_id_lookup <- tibble(
  index = 1:nrow(nodes_tbl),
  name = nodes_tbl$name
)

##
edge_df <- edges_tbl %>%
  mutate(
    from = node_id_lookup$name[from],
    to   = node_id_lookup$name[to]
  )

# Join coordinates to calculate edge lengths
edge_coords <- edge_df %>%
  left_join(transmission_layout %>% dplyr::rename(from = name, x_from = x, y_from = y), by = "from") %>%
  left_join(transmission_layout %>% dplyr::rename(to = name, x_to = x, y_to = y), by = "to") %>%
  mutate(
    edge_length = sqrt((x_to - x_from)^2 + (y_to - y_from)^2)
  )

##
# Filter long edges
threshold <- 5
short_edges <- edge_coords %>% filter(edge_length <= threshold)

## Plot
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.2,
    color = "gray70",
    linewidth = 0.4,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = cluster),
    size = 3
  ) +
  scale_color_manual(values = cluster_colors) +
  labs(title = "Muko Clusters (dcifer)", color = "Cluster ID") +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

dcifer_plot

ggsave("DCIFER_RESULTS/Muko_dcifer_Clusters_fixed.png", dcifer_plot, width = 10, height = 8, units = "in")

##----------------------------------------------
# Colour by travel history
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.5,
    color = "gray50",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = furthesttravel),
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    na.value = "gray80"
  ) +
  labs(title = "Muko Travel History", color = "Travel History") +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

dcifer_plot

ggsave("DCIFER_RESULTS/Muko_dcifer_Clusters_fixed_travel.png", dcifer_plot, width = 10, height = 8, units = "in")
##--------------------------------------------------------------------------------

### Remove the title and the legend from the Dcifer plot
# Function to extract legend
get_legend_plot <- function(plot) {
  g <- ggplotGrob(plot)
  legend_index <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  g$grobs[[legend_index]]
}

# --- Create plot WITH legend ---
dcifer_plot_with_legend <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.5,
    color = "gray50",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = furthesttravel),
    size = 6   # 🔥 Increased node size
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    na.value = "gray80",
    name = "Travel History"
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1.5, "cm"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 5))  # bigger dots in legend
  )

# --- Extract legend ---
legend_plot <- get_legend_plot(dcifer_plot_with_legend)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

ggsave("Final_Manuscript_Figures/Mu_Dcifer_legend.png",
       plot = arrangeGrob(legend_plot),
       width = 3.5, height = 3, dpi = 300, bg = "white")

# Main plot WITHOUT legend
Muko_plot_clean <- dcifer_plot_with_legend +
  theme(legend.position = "none")

Muko_plot_clean

# Save clean plot
ggsave("Final_Manuscript_Figures/Muko_dcifer_No_Legend.png",
       plot = Muko_plot_clean,
       width = 9, height = 9, units = "in", bg = "white")
```

### Build a DCIFER network plot with thickened edges based on r
```{r,}
set.seed(100)

short_edges$Pairwise_IBD <- as.numeric(short_edges$ibd)

## With no Legend
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Edges: thickness & alpha reflect IBD
  geom_segment(
    data = short_edges,   # or short_edges_sig if you filtered
    aes(
      x    = x_from,
      y    = y_from,
      xend = x_to,
      yend = y_to,
      linewidth = Pairwise_IBD,
      alpha     = Pairwise_IBD
    ),
    color = "gray30",
    show.legend = TRUE,
    inherit.aes = FALSE
  ) +
    scale_linewidth(range = c(0.5, 2)) +
  
  # Nodes: travel-history colors
  geom_node_point(
    aes(color = furthesttravel),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    na.value = "gray80",
    name = "Travel history"
  ) +
  
  coord_fixed() +
  theme_void() +
  theme(
    legend.position = "none" 
  )

dcifer_plot

ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Muko_Dcifer_Ibd_No_legend.png",
       plot = dcifer_plot,
       width = 12, height = 9, units = "in", bg = "white")
###===================================================================

##### With a standard scale
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Edges: thickness & alpha reflect IBD
  geom_segment(
    data = short_edges,   # or short_edges_sig if you filtered
    aes(
      x    = x_from,
      y    = y_from,
      xend = x_to,
      yend = y_to,
      linewidth = Pairwise_IBD,
      alpha     = Pairwise_IBD
    ),
    color = "gray30",
    show.legend = TRUE,
    inherit.aes = FALSE
  ) +
    scale_linewidth(range = c(0.5, 2)) +
  
  # Nodes: travel-history colors
  geom_node_point(
    aes(color = furthesttravel),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    na.value = "gray80",
    name = "Travel history"
  ) +
  
  coord_fixed() +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size  = unit(1.3, "cm"),
    legend.text      = element_text(size = 11),
    legend.title     = element_text(size = 13, face = "bold")
  )

dcifer_plot

ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Muko_Dcifer_Ibd_legend1.png",
       plot = dcifer_plot,
       width = 12, height = 9, units = "in", bg = "white")

##------------------------------------------------------------------------------

#### Save the legend separate
# Extract legend
legend_plot <- get_legend_plot(dcifer_plot)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Muko_dcifer_legend.png",
       plot = arrangeGrob(legend_plot),
       width = 3.5, height = 6, dpi = 300, bg = "white")
##----------------------------------------------------------------------------
#### Increase the plot in size
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Edges: thickness & alpha reflect IBD
  geom_segment(
    data = short_edges,   # or short_edges_sig if you filtered
    aes(
      x    = x_from,
      y    = y_from,
      xend = x_to,
      yend = y_to,
      linewidth = Pairwise_IBD,
      alpha     = Pairwise_IBD
    ),
    color = "gray30",
    show.legend = TRUE,
    inherit.aes = FALSE
  ) +
    scale_linewidth(range = c(0.5, 2)) +
  
  # Nodes: travel-history colors
  geom_node_point(
    aes(color = furthesttravel),
    size = 8
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    na.value = "gray80",
    name = "Travel history"
  ) +
  
  coord_fixed() +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size  = unit(1.3, "cm"),
    legend.position = "none",
    plot.margin = margin(-5, -5, -5, -5),
    legend.text      = element_text(size = 11),
    legend.title     = element_text(size = 13, face = "bold")
  )

dcifer_plot

ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Muko_Dcifer_ibd_No_Legend.png",
       plot = dcifer_plot,
       width = 12, height = 9, units = "in", bg = "white")

###==========================================================================
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Edges: thickness & alpha reflect IBD
  geom_segment(
    data = short_edges,   # or short_edges_sig if you filtered
    aes(
      x    = x_from,
      y    = y_from,
      xend = x_to,
      yend = y_to,
      linewidth = Pairwise_IBD,
      alpha     = Pairwise_IBD
    ),
    color = "gray30",
    show.legend = TRUE,
    inherit.aes = FALSE
  ) +
  scale_linewidth(range = c(0.5, 2)) +
  
  # Nodes: travel-history colors
  geom_node_point(
    aes(color = furthesttravel),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    na.value = "gray80",
    name = "Travel history"
  ) +
  
  scale_x_continuous(expand = expansion(mult = 0.04)) +
  scale_y_continuous(expand = expansion(mult = 0.04)) +

  coord_fixed() +
  theme_void() +
  theme(
    plot.margin = margin(-5, -5, -5, -5),
    legend.position = "none" 
  )

dcifer_plot


ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Muko_Dcifer_ibd_No_Legend2.png",
       plot = dcifer_plot,
       width = 8, height = 8, units = "in", bg = "white")

```
