---
title: "Dcifer Analysis"
author: "Monica Mbabazi"
date: "2025-08-19th"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls(all=TRUE))
## Loading required libraries
library(dcifer)
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(readr)
library(plotly)
library(ggnewscale)
library(tidyverse)
library(descriptr)
library(ggraph)
library(igraph)
library(grid)
library(kableExtra)
library(cluster)  # For kmeans clustering
library(mclust)  
library(igraph)
library(scales)
library(tidygraph)
library(gridExtra)
setwd("/home/monica/Desktop/IMMRSE_Study/IMMRSE-Concept/Manuscript_Analysis_03_2025/")
```

#### Running Dcifer on Kamwezi Data
```{r, echo=TRUE}
## Moire outputs
Kamwezi_moire_results <- readr::read_rds("MOIRE_RESULTS/Kamwezi/Kamwezi_mcmc_results.rds")

## Moire data
Kamwezi_data_moire <- read.csv("Filtered_data/Kamwezi_filtered_data.csv", header = T)
#Kamwezi_data_moire <- Kamwezi_data_moire %>% rename("allele"="PseudoCIGAR")
# Format the data
Kamwezi_data <- moire::load_long_form_data(Kamwezi_data_moire)

# Function to calculate median allele frequencies
calculate_median_afs <- function(x) {
  median_afs <- lapply(x$chains[[1]]$allele_freqs, function(af) {
    mat <- matrix(unlist(af), ncol = length(af))
    d <- dist(t(mat))
    af[[which.min(rowSums(as.matrix(d)))]]
  })
  return(median_afs)
}
# Calculate median allele frequencies 
Kamwezi_median_afs <- calculate_median_afs(Kamwezi_moire_results)
# Define uninformative loci
ka_uninformative_loci <- Kamwezi_data$uninformative_loci

# Filter to remove these loci
filtered_Kamwezi <- Kamwezi_data_moire %>% filter(!locus %in% ka_uninformative_loci)

# Prepare DCIFER input for each site
Kamwezi_dsmp<- dcifer::formatDat(filtered_Kamwezi, "sample_id", "locus", "allele")
### COI
Kamwezi_coi   <- moire::summarize_coi(Kamwezi_moire_results)$post_coi_med

### Median frequency
Kamwezi_dsmp <- dcifer::matchAfreq(Kamwezi_dsmp, Kamwezi_median_afs)

## IBD per site
dres_df <-ibdDat(Kamwezi_dsmp, Kamwezi_coi, Kamwezi_median_afs, pval = TRUE, confint = TRUE, rnull = 0.125, alpha = 0.05, nr = 1e3)

```

### Highly related pairs
```{r,}
### Visualize the distribution of IBD estimates
# Extract IBD estimates from rnull = 0.25 run
ibd_est <- dres_df[, , "estimate"]
ibd_lower <- dres_df[, , "CI_lower"]
ibd_upper <- dres_df[, , "CI_upper"]


#do stupid stuff to make dataframe (there's probably a better way to to this)
#rnull = 0.125
ibd_df <- as.data.frame(ibd_est)
ibd_df <- tibble::rownames_to_column(ibd_df, var = "x")

ibd_lower_df <- as.data.frame(ibd_lower)
ibd_lower_df <- tibble::rownames_to_column(ibd_lower_df, var = "x")

ibd_upper_df <- as.data.frame(ibd_upper)
ibd_upper_df <- tibble::rownames_to_column(ibd_upper_df, var = "x")

#this is a DF with all ibd data in longform between non-matching pairs of samples
ibd_df_long <- ibd_df %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "ibd"
  ) %>% filter(!is.na(ibd)) 

hist(ibd_df_long$ibd)

#### BIND IN CI VALUES TO EACH DF ###############

CI_lower_df_long <- ibd_lower_df %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "ibd_lower")

CI_upper_df_long <- ibd_upper_df %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "ibd_upper")

#bind CIs into long df 
ibd_df_long <- ibd_df_long %>% 
  left_join(CI_lower_df_long) %>% 
  left_join(CI_upper_df_long)

#### BIND IN P VALUES TO EACH DF ###############
pvals <- dres_df[, , "p_value"]
pvals <- as.data.frame(pvals)
pvals <- tibble::rownames_to_column(pvals, var = "x")

#pval 0.05
pvals_df_long <- pvals %>%
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "p_value"
  ) %>% filter(!is.na(p_value)) %>% 
  mutate(p_value = p_value/2) #divide by 2 for 1 sided test

#bind pvals into long df 
ibd_df_long <- ibd_df_long %>% 
  left_join(pvals_df_long)


#make p-value very large if r <0.125

ibd_df_long <- ibd_df_long %>% mutate(p_value = ifelse(ibd <0.125, 1, p_value))

#here is where i do corrections for rnull = 0.25 dataframe
long_ibd1 <- ibd_df_long %>% 
  mutate(p_BH = p.adjust(p_value, method = "BH"), 
         p_BF = p.adjust(p_value, method = "bonferroni")) %>% 
  mutate(sig_0.01 = ifelse(p_value < 0.01, 1, 0), sig_0.01_BH = ifelse(p_BH < 0.01, 1, 0), 
         sig_0.01_BF = ifelse(p_BF < 0.01, 1, 0)) %>% 
  mutate(sig_0.05 = ifelse(p_value < 0.05, 1, 0), sig_0.05_BH = ifelse(p_BH < 0.05, 1, 0),
         sig_0.05_BF= ifelse(p_BF < 0.05, 1, 0)) %>% 
  mutate(n = 1)

#highly related pairs? 
highly_related_df <- long_ibd1 %>% filter(sig_0.05_BH == 1)
write.csv(highly_related_df, 
          file = "Final_Manuscript_Figures/Re-analysis_Figs_2026/Kamwezi_highly_rel-pairs.csv", 
          row.names = FALSE)
#highly related pairs out of all pairwise comparisons # 0.06363636 (6.36%)
sum(long_ibd1$sig_0.05_BH)/sum(long_ibd1$n)

## r ranges
range(highly_related_df$ibd, na.rm = TRUE) # 0.294 - 1.000
```

### Network plots based on DCIFER
```{r,}
# Extract IBD estimates for each site
dmat_Kamwezi <- dres_df[, , "estimate"]
## Plotting: Include also the singletons in the plot
g <- graph_from_data_frame(highly_related_df, vertices = data.frame(name = rownames(dmat_Kamwezi)), directed = FALSE)

# Cluster detection
clusters <- components(g)

# Reassign singleton clusters to "Singleton"
cluster_sizes <- table(clusters$membership)
singleton_clusters <- names(cluster_sizes[cluster_sizes == 1])
cluster_labels <- as.character(clusters$membership)
cluster_labels[cluster_labels %in% singleton_clusters] <- "Singleton"

# Relabel non-singleton clusters in descending size order (so biggest is "1")
multi_cluster_sizes <- sort(table(cluster_labels[cluster_labels != "Singleton"]), decreasing = TRUE)
new_ids <- setNames(seq_along(multi_cluster_sizes), names(multi_cluster_sizes))
cluster_labels <- ifelse(cluster_labels == "Singleton", "Singleton", as.character(new_ids[cluster_labels]))

# Build tidygraph
tg <- as_tbl_graph(g) %>%
  mutate(cluster = factor(cluster_labels, levels = c("Singleton", as.character(seq_along(multi_cluster_sizes)))))

# Set colors: gray for singleton, rainbow for others
cluster_colors <- setNames(
  c("gray", rainbow(length(multi_cluster_sizes))),
  levels(tg$cluster)
)

##
# Plot
p <- ggraph(tg, layout = "fr") +
  geom_edge_link(alpha = 0.3) +
  geom_node_point(aes(color = cluster), size = 3) +
  scale_color_manual(values = cluster_colors) +
  labs(title = "Clusters of Highly Related Samples (Kamwezi)",
       color = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )

p

```

## Transmission Network Plots
```{r, }
### Resistance data
#subset these barcodes for the barcodes in our timeframe 
Res_genotypes <- read_csv("Resistance_Analysis/Resistance_Data/sample_genotypes_all.csv")

# Keep only samples from Kamwezi
Res_genotypes <- Res_genotypes %>% 
  filter(Sample.Site == "Kamwezi") %>% dplyr::rename(sample_id=barcode)

##### TABULATE Ns #####
##### TABULATE Ns #####
meta_long <- pivot_longer(Res_genotypes, cols = arps10_V127M:k13_C469F, names_to = "SNP", values_to = "Genotype") 

#k13 only
meta_long <- meta_long %>% filter(grepl("k13", SNP))

##
k13_P441L <- meta_long %>% filter(SNP == "k13_P441L") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_P441L = Genotype)

#
k13_675 <- meta_long %>% filter(SNP == "k13_A675V") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_675 = Genotype)

k13_C469Y <- meta_long %>% filter(SNP == "k13_C469Y") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_C469Y = Genotype)

k13_C469F <- meta_long %>% filter(SNP == "k13_C469F") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_C469F = Genotype)

k13_R561H <- meta_long %>% filter(SNP == "k13_R561H") %>% 
  dplyr::select(sample_id, Genotype) %>% 
  dplyr::rename(id = sample_id) %>% 
  mutate(Genotype = ifelse(Genotype == 1 | Genotype == 2, "Mix or Mutant", "WT")) %>% 
  dplyr::rename(k13_R561H = Genotype)

#### Survey Data
library(haven)

zumba_travel <- read_dta("Data/Health_Facility_Survey_Kamwezi_24-01-2025.dta")
names(zumba_travel)

travel2 <- zumba_travel %>% dplyr::select(fpbarcode, furthest_travel, starttime, firstdate1, lastdate1) %>% 
  dplyr::rename(id = fpbarcode, date = starttime)

travel2 <- travel2 %>% mutate(furthesttravel = case_when(furthest_travel == 1 ~ "No travel", 
                                                         furthest_travel >1 & furthest_travel <=4 ~ "Inside district", 
                                                         furthest_travel == 5 ~ "Outside district inside country", 
                                                         furthest_travel == 6 ~ "Outside country", 
                                                         TRUE ~ "Missing data")) 
table(travel2$furthesttravel)

travel2 <- travel2 %>% dplyr::select(-furthest_travel)

travel2 <- travel2 %>%
  mutate(furthesttravel = ifelse(is.na(furthesttravel), "Missing data", furthesttravel)) %>%
  mutate(furthesttravel = factor(furthesttravel, levels = c(
    "Outside district inside country", 
    "Inside district", 
    "No travel", 
    "Missing data"
  ), ordered = TRUE))

## Save this metadata as a csv file
write.csv(travel2, "Data/travel_subset_cleaned_Kamw.csv", row.names = FALSE)
###------------------------------------------------------------------------------------
#### Transmission Network data
kamwezi_cs <- readRDS("Transmission_Network_Analysis/kamwezi_inferred_network_edges_cs.rds")

edge_output <- kamwezi_cs %>%
  dplyr::rename(from = parent_node, to = sample_id, prob = mean_posterior_prob)

# Remove edges from nodes with `_copy` as from — we don't want `_copy` nodes in the plot
edge_output_clean <- edge_output %>%
  filter(!str_detect(from, "_copy$"))
#
real_node_ids <- unique(c(edge_output$to, edge_output_clean$from))
#
founder_nodes <- setdiff(edge_output_clean$from, edge_output_clean$to)
## Observed founders (no incoming edge)
observed_founders <- setdiff(edge_output_clean$from, edge_output_clean$to)

## Singleton founders (only infected by _copy and never infected by anyone else)
singleton_founders <- edge_output %>%
  filter(str_detect(from, "_copy$")) %>%
  pull(to) %>%
  setdiff(edge_output_clean$to)

# Combine both groups
all_founders <- union(observed_founders, singleton_founders)
#
graph_nodes_annotated <- tibble(name = real_node_ids) %>%
  left_join(travel2, by = c("name" = "id")) %>%
  left_join(k13_P441L,         by = c("name" = "id")) %>%
  left_join(k13_675,       by = c("name" = "id")) %>%
  left_join(k13_C469Y,     by = c("name" = "id")) %>%
  left_join(k13_C469F,     by = c("name" = "id")) %>%
  left_join(k13_R561H,     by = c("name" = "id")) %>%
  mutate(
    founder = name %in% all_founders
  )
#### ADD COI values 

coi_df <- read.csv("MOIRE_RESULTS/Kamwezi/coi_summary.csv", header = TRUE) %>% rename(name = sample_id) 
## 
graph_nodes_annotated <- graph_nodes_annotated %>% 
  left_join( coi_df %>% select(name, post_coi_med), by = "name" ) %>% 
  rename(coi = post_coi_med)

# Create final transmission graph
final_graph <- graph_from_data_frame(
  d = edge_output_clean,
  vertices = graph_nodes_annotated,
  directed = TRUE
)
##===================================================================

## Plot
kamwezi_travel_plot <- ggraph(final_graph, layout = 'fr') +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Furthest Travel"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  theme_void() +
  guides(size = "none") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  ggtitle("Kamwezi Transmission Network Colored by Travel")

kamwezi_travel_plot

```

## Transmission Network Plot with fixed clusters
```{r,}
set.seed(100)
## Generate layout positions using one graph (say dcifer)
shared_layout <- create_layout(final_graph, layout = "fr")
#
shared_layout <- shared_layout %>%
  mutate(
    furthesttravel = factor(
      furthesttravel,
      levels = c(
        "Outside district inside country",
        "Inside district",
        "No travel",
        "Missing data"
      )
    )
  )

## Transmission Network Plot
kamwezi_travel_plot <- ggraph(shared_layout) +  # use shared_layout here
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "Outside the country" = "#B22222",
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Travel History"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  theme_void() +
  guides(size = "none") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  ggtitle("Kamwezi Transmission Network")

kamwezi_travel_plot

ggsave("DCIFER_RESULTS/Kamwezi_TN_Clusters_fixed_travel.png", kamwezi_travel_plot, width = 10, height = 8, units = "in")
######--------------------------------------------------------------------------

### By P441L
kamwezi_mutation_plot <- ggraph(shared_layout) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = k13_P441L, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "WT" = "#1b9e77",
      "Mix or Mutant" = "#d95f02"
    ),
    na.value = "gray80",
    name = "k13 P441L"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  theme_void() +
  guides(size = "none") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  ggtitle("Kamwezi Transmission Network by k13_P441L")

kamwezi_mutation_plot

ggsave("DCIFER_RESULTS/Kamwezi_TN_Clusters_fixed_P441L.png", kamwezi_mutation_plot, width = 10, height = 8, units = "in")
#---------------------------------------------------------------------
## R561H
kamwezi_mutation_plot <- ggraph(shared_layout) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = k13_R561H, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "WT" = "#1b9e77",
      "Mix or Mutant" = "#d95f02"
    ),
    na.value = "gray80",
    name = "k13_R561H"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 3)) +
  theme_void() +
  guides(size = "none") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  ggtitle("Kamwezi Transmission Network by k13_R561H")

kamwezi_mutation_plot

ggsave("DCIFER_RESULTS/Kamwezi_TN_Clusters_fixed_R561H.png", kamwezi_mutation_plot, width = 10, height = 8, units = "in")
##----------------------------------------------------------------------------------------

#### ### Extract the legend and remove it from the network plot
# Function to extract legend
get_legend_plot <- function(plot) {
  g <- ggplotGrob(plot)
  legend_index <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  g$grobs[[legend_index]]
}

# Plot WITH only two legends
Kamwezi_plot_with_legend <- ggraph(shared_layout) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = furthesttravel, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    name = "Travel History"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 6, `TRUE` = 6), guide = "none") +  # <-- no size legend
  theme_void() +
  theme(
    legend.key.size = unit(1.5, "cm"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 5)),
    shape = guide_legend(override.aes = list(size = 5))
  )

# Extract legend
legend_plot <- get_legend_plot(Kamwezi_plot_with_legend)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

#ggsave("Final_Manuscript_Figures/Mu_Transmission_legend.png",
 #      plot = arrangeGrob(legend_plot),
  #     width = 3.5, height = 4.5, dpi = 300, bg = "white")

# Main plot WITHOUT legend
Kamwezi_plot_clean <- Kamwezi_plot_with_legend +
  theme(legend.position = "none")

Kamwezi_plot_clean
ggsave("Final_Manuscript_Figures/Kamwezi_TN_Clusters_No_Legend.png",Kamwezi_plot_clean, width = 9, height = 9, units = "in", bg="white")
##--------------------------------------------------------------------------------------------
```
### Transmission networks with probability-weighted edges
```{r, }
set.seed(100)

Kamwezi_trans_clean <- ggraph(shared_layout) +
  geom_edge_link(
    aes(alpha = prob, width = prob),
    arrow   = arrow(type = "closed", length = unit(2, "mm")),
    end_cap = circle(3, "mm"),
    colour  = "black",
    show.legend = TRUE
  ) +

  # Edge legend scales: fixed 0–1 range, same breaks for all sites
  scale_edge_alpha(
    range  = c(0.25, 1),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +
  scale_edge_width(
    range  = c(0.4, 2),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +

  # Nodes
  geom_node_point(
    aes(color = furthesttravel, shape = founder),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    name = "Travel history"
  ) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 17),
    name   = "Founder"
  ) +

  guides(
    edge_width = guide_legend(order = 1),
    edge_alpha = guide_legend(order = 1),
    color      = guide_legend(order = 2),
    shape      = guide_legend(order = 3)
  ) +
  theme_void() +
 theme(
    legend.position = "none"
  )

Kamwezi_trans_clean

#ggsave("Final_Manuscript_Figures/Kamwezi_TN_Clusters_trans_No_legend.png",
      # Kamwezi_trans_clean, width = 10, height = 9, units = "in", bg = "white")
##----------------------------------------------------------------------------------------

Kamwezi_trans_clean <- ggraph(shared_layout) +
  geom_edge_link(
    aes(alpha = prob, width = prob),
    arrow   = arrow(type = "closed", length = unit(2, "mm")),
    end_cap = circle(3, "mm"),
    colour  = "black",
    show.legend = TRUE
  ) +

  # Edge legend scales: fixed 0–1 range, same breaks for all sites
  scale_edge_alpha(
    range  = c(0.25, 1),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +
  scale_edge_width(
    range  = c(0.4, 2),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1.0),
    name   = "Transmission\nprobability"
  ) +

  # Nodes
  geom_node_point(
    aes(color = furthesttravel, shape = founder),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    name = "Travel history"
  ) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 17),
    name   = "Founder"
  ) +

  guides(
    edge_width = guide_legend(order = 1),
    edge_alpha = guide_legend(order = 1),
    color      = guide_legend(order = 2),
    shape      = guide_legend(order = 3)
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title    = element_text(size = 12, face = "bold"),
    legend.text     = element_text(size = 11)
  )

Kamwezi_trans_clean

#ggsave("Final_Manuscript_Figures/Kamwezi_TN_Clusters_trans_legend.png",
 #      Kamwezi_trans_clean, width = 12, height = 9, units = "in", bg = "white")
##------------------------------------------------------------------------------

#### Save the legend separate
# Extract legend
legend_plot <- get_legend_plot(Kamwezi_trans_clean)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

#ggsave("Final_Manuscript_Figures/Kamwezi_trans_clean_legend.png",
 #      plot = arrangeGrob(legend_plot),
  #     width = 3.5, height = 4.5, dpi = 300, bg = "white")

```


### Exclude the single nodes (nodes with no edges): Only Clusters and color them by Mutations
```{r, }
set.seed(100)
## P441L
# 1) Create a single, fixed layout on the FULL graph
layout_full <- create_layout(final_graph, layout = "fr") %>%
  as_tibble() %>%
  select(name, x, y)

# 2) Compute degree on the FULL graph (undirected degree for “singleton” test)
deg_vec <- igraph::degree(final_graph, mode = "all")
degree_df <- tibble(name = V(final_graph)$name, degree = as.numeric(deg_vec))

# 3) Join coordinates + degree; pick the nodes to KEEP (degree > 0)
node_pos <- layout_full %>%
  left_join(degree_df, by = "name")

keep_nodes <- node_pos %>%
  filter(degree > 0) %>%
  pull(name)

# 4) Induce a subgraph of ONLY connected nodes (no singletons), but KEEP attributes
g_clusters <- induced_subgraph(final_graph, vids = V(final_graph)[name %in% keep_nodes])

# 5) Prepare manual coordinates for the subgraph (match to full layout)
coord_map <- node_pos %>%
  filter(name %in% keep_nodes) %>%
  select(name, x, y)

idx <- match(V(g_clusters)$name, coord_map$name)
manual_x <- coord_map$x[idx]
manual_y <- coord_map$y[idx]

# 6) Clean up k13_P441L labels (NAs -> "Missing data") & set fixed order
vt <- vertex_attr(g_clusters)
k13 <- vt$k13_P441L
k13[is.na(k13)] <- "Missing data"
k13 <- factor(
  k13,
  levels = c("WT", "Mix or Mutant", "Missing data"),
  ordered = TRUE
)
g_clusters <- set_vertex_attr(g_clusters, "k13_P441L", value = k13)

# 7) Plot with MANUAL layout (positions are fixed), edges & alpha from 'prob'
#    Color by k13_P441L; shape by founder; hide size legend.
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = k13_P441L, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "WT" = "#1b9e77",
      "Mix or Mutant" = "#d95f02"
    ),
    name = "k13_P441L"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 6, `TRUE` = 6)) +
  guides(size = "none") +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1.5, "cm"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold")
  )+
  guides(
    color = guide_legend(override.aes = list(size = 5)),
    shape = guide_legend(override.aes = list(size = 5))
  )

print(p)

#ggsave("Final_Manuscript_Figures/Kamwezi_P441L.png", p, width = 10, height = 8, units = "in")

### 
#### Save the legend separate
# Extract legend
legend_plot <- get_legend_plot(p)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

ggsave("Final_Manuscript_Figures/Kamwezi_P441L_legend.png",
       plot = arrangeGrob(legend_plot),
       width = 3, height = 4, dpi = 300, bg = "white")
##--------------------------------------------------------------------------------------------
## Remove the legend from the figure
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = k13_P441L, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "WT" = "#1b9e77",
      "Mix or Mutant" = "#d95f02"
    ),
    name = "k13_P441L"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 6, `TRUE` = 6)) +
  guides(size = "none") +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1.5, "cm"),
    legend.text = element_text(size = 12),
    legend.position = "none",
    legend.title = element_text(size = 14, face = "bold")
  )+
  guides(
    color = guide_legend(override.aes = list(size = 5)),
    shape = guide_legend(override.aes = list(size = 5))
  )

print(p)

#ggsave("Final_Manuscript_Figures/Kamwezi_P441L_No_Legend.png", p, width = 10, height = 8, units = "in")
###=================================================================================

# R561H

# 6) Clean up k13_R561H labels (NAs -> "Missing data") & set fixed order
vt <- vertex_attr(g_clusters)
k13 <- vt$k13_R561H
k13[is.na(k13)] <- "Missing data"
k13 <- factor(
  k13,
  levels = c("WT", "Mix or Mutant", "Missing data"),
  ordered = TRUE
)
g_clusters <- set_vertex_attr(g_clusters, "k13_R561H", value = k13)

# 7) Plot with MANUAL layout (positions are fixed), edges & alpha from 'prob'
#    Color by k13_R561H; shape by founder; hide size legend.
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(aes(edge_alpha = prob),
                 arrow = arrow(type = "closed", length = unit(2, 'mm')),
                 end_cap = circle(2, 'mm'),
                 show.legend = FALSE) +
  geom_node_point(aes(color = k13_R561H, shape = founder, size = founder)) +
  scale_color_manual(
    values = c(
      "WT" = "#1b9e77",
      "Mix or Mutant" = "#d95f02"
    ),
    name = "k13_R561H"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Founder") +
  scale_size_manual(values = c(`FALSE` = 6, `TRUE` = 6)) +
  guides(size = "none") +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1.5, "cm"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold")
  )+
  guides(
    color = guide_legend(override.aes = list(size = 5)),
    shape = guide_legend(override.aes = list(size = 5))
  )

print(p)

#ggsave("Final_Manuscript_Figures/Kamwezi_R561H.png", p, width = 10, height = 8, units = "in")

##----------------- COI -------------------------------------------------------------

# Extract raw COI values
vt  <- vertex_attr(g_clusters)
coi_raw <- vt$coi

# Build two COI categories
coi_cat <- dplyr::case_when(
  is.na(coi_raw) ~ "Missing",
  coi_raw == 1   ~ "Monoclonal",
  coi_raw > 1    ~ "Polyclonal"
)

coi_cat <- factor(coi_cat, levels = c("Monoclonal", "Polyclonal", "Missing"))

# Add to graph
g_clusters <- set_vertex_attr(g_clusters, "COI_cat", value = coi_cat)

## Plot
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(
    aes(edge_alpha = prob),
    arrow = arrow(type = "closed", length = unit(2, 'mm')),
    end_cap = circle(2, 'mm'),
    show.legend = FALSE
  ) +
  
  # Node aesthetics
  geom_node_point(
    aes(color = k13_P441L, shape = COI_cat),
    size = 6
  ) +
  
  # Color scale: K13 genotype
  scale_color_manual(
    values = c(
      "WT"            = "#1b9e77",
      "Mix or Mutant" = "#d95f02",
      "Missing data"  = "#969696"
    ),
    name = "Genotype"
  ) +
  
  # Shape scale: COI categories
  scale_shape_manual(
    values = c(
      "Monoclonal" = 18,  # diamond
      "Polyclonal" = 16,  # solid circle
      "Missing"    = 1    # hollow circle
    ),
    name   = "COI"
  ) +
  
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size  = unit(1.5, "cm"),
    legend.text      = element_text(size = 12),
    legend.title     = element_text(size = 14, face = "bold")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 5)),
    shape = guide_legend(override.aes = list(size = 5))
  )

print(p)
#ggsave("Final_Manuscript_Figures/Figure 3/Kamwezi_P441L_COI.png", p, width = 10, height = 8, units = "in")
#### Save the legend separate
# Extract legend
legend_plot <- get_legend_plot(p)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

#ggsave("Final_Manuscript_Figures/Figure 3/Kamwezi_P441L_legend.png",
 #      plot = arrangeGrob(legend_plot),
  #     width = 3, height = 4, dpi = 300, bg = "white")

##---------------------------------------------------------------------------------
#### some changes on the plot
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(
    aes(edge_alpha = prob),
    arrow = arrow(type = "closed", length = unit(2, 'mm')),
    end_cap = circle(2, 'mm'),
    show.legend = FALSE
  ) +
  
  # Larger nodes
  geom_node_point(
    aes(color = k13_P441L, shape = COI_cat),
    size = 8
  ) +
  
  # Color scale
  scale_color_manual(
    values = c(
      "WT"            = "#1b9e77",
      "Mix or Mutant" = "#d95f02",
      "Missing data"  = "#969696"
    ),
    name = "Genotype"
  ) +
  
  # Shape scale
  scale_shape_manual(
    values = c(
      "Monoclonal" = 18,  # diamond
      "Polyclonal" = 16,  # solid circle
      "Missing"    = 1    # hollow circle
    ),
    name = "COI"
  ) +
  
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),

    # Tighter legend, less whitespace
    legend.key.height = unit(1.5, "cm"),
    legend.key.width  = unit(1, "cm"),
    legend.spacing.y  = unit(0.4, "cm"),
    legend.spacing.x  = unit(0.4, "cm"),
    
    # Larger legend text
    legend.text  = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold")
  ) +
  
  # Increase size of legend symbols
  guides(
    color = guide_legend(override.aes = list(size = 7)),
    shape = guide_legend(override.aes = list(size = 7))
  )

print(p)

#ggsave("Final_Manuscript_Figures/Figure 3/Kamwezi_P441L_COI_bigger.png", p, width = 10, height = 8, units = "in")


#### Save the legend separate
# Extract legend
legend_plot <- get_legend_plot(p)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

#ggsave("Final_Manuscript_Figures/Figure 3/Kamwezi_P441L_legend_bigger.png",
 #      plot = arrangeGrob(legend_plot),
  #     width = 2.5, height = 3, dpi = 300, bg = "white")
##--------------------------------------------------------------------------------------------
## Remove the legend from the figure
### No Legend plot

p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(
    aes(edge_alpha = prob),
    arrow = arrow(type = "closed", length = unit(2, 'mm')),
    end_cap = circle(2, 'mm'),
    show.legend = FALSE
  ) +
  
  # Larger nodes
  geom_node_point(
    aes(color = k13_P441L, shape = COI_cat),
    size = 8
  ) +
  
  # Color scale
  scale_color_manual(
    values = c(
      "WT"            = "#1b9e77",
      "Mix or Mutant" = "#d95f02",
      "Missing data"  = "#969696"
    ),
    name = "Genotype"
  ) +
  
  # Shape scale
  scale_shape_manual(
    values = c(
      "Monoclonal" = 18,  # diamond
      "Polyclonal" = 16,  # solid circle
      "Missing"    = 1    # hollow circle
    ),
    name = "COI"
  ) +
  
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    legend.position = "none",
    panel.background = element_rect(fill = "white", color = NA),

    # Tighter legend, less whitespace
    legend.key.height = unit(1.5, "cm"),
    legend.key.width  = unit(1, "cm"),
    legend.spacing.y  = unit(0.4, "cm"),
    legend.spacing.x  = unit(0.4, "cm"),
    
    # Larger legend text
    legend.text  = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold")
  ) +
  
  # Increase size of legend symbols
  guides(
    color = guide_legend(override.aes = list(size = 7)),
    shape = guide_legend(override.aes = list(size = 7))
  )

print(p)

#ggsave("Final_Manuscript_Figures/Figure 3/Kamwezi_P441L_COI_No_Legend.png", p, width = 10, height = 8, units = "in")
###================================================================================================

## K13 R561H
#### some changes on the plot
p <- ggraph(g_clusters, layout = "manual", x = manual_x, y = manual_y) +
  geom_edge_link(
    aes(edge_alpha = prob),
    arrow = arrow(type = "closed", length = unit(2, 'mm')),
    end_cap = circle(2, 'mm'),
    show.legend = FALSE
  ) +
  
  # Larger nodes
  geom_node_point(
    aes(color = k13_R561H, shape = COI_cat),
    size = 8
  ) +
  
  # Color scale
  scale_color_manual(
    values = c(
      "WT"            = "#1b9e77",
      "Mix or Mutant" = "#d95f02",
      "Missing data"  = "#969696"
    ),
    name = "Genotype"
  ) +
  
  # Shape scale
  scale_shape_manual(
    values = c(
      "Monoclonal" = 18,  # diamond
      "Polyclonal" = 16,  # solid circle
      "Missing"    = 1    # hollow circle
    ),
    name = "COI"
  ) +
  
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    legend.position = "none",
    panel.background = element_rect(fill = "white", color = NA),

    # Tighter legend, less whitespace
    legend.key.height = unit(1.5, "cm"),
    legend.key.width  = unit(1, "cm"),
    legend.spacing.y  = unit(0.4, "cm"),
    legend.spacing.x  = unit(0.4, "cm"),
    
    # Larger legend text
    legend.text  = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold")
  ) +
  
  # Increase size of legend symbols
  guides(
    color = guide_legend(override.aes = list(size = 7)),
    shape = guide_legend(override.aes = list(size = 7))
  )

print(p)

#ggsave("Final_Manuscript_Figures/Figure 3/Kamwezi_R561H_COI_No_Legend.png", p, width = 10, height = 8, units = "in")
```



## DCIFER Network Plot with fixed clusters
```{r, }
set.seed(100)
# Extract node positions from shared layout

#Extract node name and x/y
transmission_layout <- shared_layout %>% 
  dplyr::select(name, x, y)

##
tg_with_coords <- tg %>%
  activate(nodes) %>%
  left_join(transmission_layout, by = "name")

## Plot
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Transparent, thin, gray edges
  geom_edge_link(
    alpha = 0.2,
    color = "gray70",
    edge_width = 0.4
  ) +
  # Optional: Use curved edges to reduce clutter
  # geom_edge_fan(alpha = 0.2, color = "gray70", edge_width = 0.3)

  # Plot nodes by cluster
  geom_node_point(
    aes(color = cluster),
    size = 3
  ) +
  scale_color_manual(values = cluster_colors) +  # Make sure this exists
  labs(title = "Kamwezi Clusters (dcifer)", color = "Cluster ID") +
  coord_fixed() +
  # Clean white background
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

dcifer_plot

###-------------------------------------------------------
## Excluding long edges from the plot
# Extract layout (should be character node names)
transmission_layout <- shared_layout %>% dplyr::select(name, x, y)

#
tg_with_coords <- tg %>%
  activate(nodes) %>%
  left_join(transmission_layout, by = "name")

## Add travel info
nodes_tbl <- tg_with_coords %>%
  activate(nodes) %>%
  as_tibble() %>%
  left_join(travel2, by = c("name" = "id"))  

##
tg_with_coords <- tg_with_coords %>%
  activate(nodes) %>%
  left_join(travel2, by = c("name" = "id"))

 # Extract edge table and node table
edges_tbl <- tg_with_coords %>%
  activate(edges) %>%
  as_tibble()

#
nodes_tbl <- tg_with_coords %>%
  activate(nodes) %>%
  as_tibble() %>%
  mutate(row_id = row_number())
#
node_id_lookup <- tibble(
  index = 1:nrow(nodes_tbl),
  name = nodes_tbl$name
)

##
edge_df <- edges_tbl %>%
  mutate(
    from = node_id_lookup$name[from],
    to   = node_id_lookup$name[to]
  )

# Join coordinates to calculate edge lengths
edge_coords <- edge_df %>%
  left_join(transmission_layout %>% dplyr::rename(from = name, x_from = x, y_from = y), by = "from") %>%
  left_join(transmission_layout %>% dplyr::rename(to = name, x_to = x, y_to = y), by = "to") %>%
  mutate(
    edge_length = sqrt((x_to - x_from)^2 + (y_to - y_from)^2)
  )

# Filter long edges
threshold <- 5
short_edges <- edge_coords %>% filter(edge_length <= threshold)

## Plot
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.4,
    color = "gray70",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = cluster),
    size = 3
  ) +
  scale_color_manual(values = cluster_colors) +
  labs(title = "Kamwezi Clusters (dcifer)", color = "Cluster ID") +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

dcifer_plot

ggsave("DCIFER_RESULTS/Kamwezi_dcifer_Clusters_fixed.png", dcifer_plot, width = 10, height = 8, units = "in")

####------------------------------------------------------------------
# Color by travel history
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.5,
    color = "gray50",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = furthesttravel),
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "Outside the country" = "#B22222",
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    na.value = "gray80"
  ) +
  labs(title = "Kamwezi Travel History", color = "Travel History") +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

dcifer_plot

ggsave("DCIFER_RESULTS/Kamwezi_dcifer_Clusters_fixed_travel.png", dcifer_plot, width = 10, height = 8, units = "in")
##------------------------------------------------------------------------------

### Remove the title and the legend from the Dcifer plot
# Function to extract legend
get_legend_plot <- function(plot) {
  g <- ggplotGrob(plot)
  legend_index <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  g$grobs[[legend_index]]
}

# --- Create plot WITH legend ---
dcifer_plot_with_legend <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.5,
    color = "gray50",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = furthesttravel),
    size = 6   # 🔥 Increased node size
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district" = "#359B73",
      "No travel" = "#F748A5",
      "Missing data" = "#969696"
    ),
    na.value = "gray80",
    name = "Travel History"
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1.5, "cm"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 5))  # bigger dots in legend
  )

# --- Extract legend ---
legend_plot <- get_legend_plot(dcifer_plot_with_legend)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

ggsave("Final_Manuscript_Figures/Ka_Dcifer_legend.png",
       plot = arrangeGrob(legend_plot),
       width = 3.5, height = 3, dpi = 300, bg = "white")

# Main plot WITHOUT legend
Kamwezi_plot_clean <- dcifer_plot_with_legend +
  theme(legend.position = "none")

Kamwezi_plot_clean

# Save clean plot
ggsave("Final_Manuscript_Figures/Kamwezi_dcifer_No_Legend.png",
       plot = Kamwezi_plot_clean,
       width = 9, height = 9, units = "in", bg = "white")

```

### Build a DCIFER network plot with thickened edges based on r
```{r,}
set.seed(100)

short_edges$Pairwise_IBD <- as.numeric(short_edges$ibd)

## With no Legend
dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Edges: thickness & alpha reflect IBD
  geom_segment(
    data = short_edges,   # or short_edges_sig if you filtered
    aes(
      x    = x_from,
      y    = y_from,
      xend = x_to,
      yend = y_to,
      linewidth = Pairwise_IBD,
      alpha     = Pairwise_IBD
    ),
    color = "gray30",
    show.legend = TRUE,
    inherit.aes = FALSE
  ) +
  scale_linewidth(range = c(0.5, 2)) +
  
  # Nodes: travel-history colors
  geom_node_point(
    aes(color = furthesttravel),
    size = 8
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    na.value = "gray80",
    name = "Travel history"
  ) +
  
  scale_x_continuous(expand = expansion(mult = 0.04)) +
  scale_y_continuous(expand = expansion(mult = 0.04)) +
  
  coord_fixed() +
  theme_void() +
  theme(
    plot.margin = margin(-5, -5, -5, -5),
    legend.position = "none" 
  )

dcifer_plot

ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Kamwez1_Dcifer_Ibd_No_legend1.png",
       plot = dcifer_plot,
       width = 12, height = 9, units = "in", bg = "white")
##
ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Kamwez1_Dcifer_Ibd_No_legend2.png",
       plot = dcifer_plot,
       width = 8, height = 8, units = "in", bg = "white")
##==================================================================================

dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  # Edges: thickness & alpha reflect IBD
  geom_segment(
    data = short_edges,   # or short_edges_sig if you filtered
    aes(
      x    = x_from,
      y    = y_from,
      xend = x_to,
      yend = y_to,
      linewidth = Pairwise_IBD,
      alpha     = Pairwise_IBD
    ),
    color = "gray30",
    show.legend = TRUE,
    inherit.aes = FALSE
  ) +
  scale_linewidth(range = c(0.5, 2)) +
  
  # Nodes: travel-history colors
  geom_node_point(
    aes(color = furthesttravel),
    size = 6
  ) +
  scale_color_manual(
    values = c(
      "Outside district inside country" = "#2271B2",
      "Inside district"                = "#359B73",
      "No travel"                      = "#F748A5",
      "Missing data"                   = "#969696"
    ),
    na.value = "gray80",
    name = "Travel history"
  ) +

  coord_fixed() +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key.size  = unit(1.3, "cm"),
    legend.text      = element_text(size = 11),
    legend.title     = element_text(size = 13, face = "bold")
  )

dcifer_plot
ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Kamwezi_Dcifer_Ibd_legend.png",
       plot = dcifer_plot,
       width = 12, height = 9, units = "in", bg = "white")

#### Save the legend separate
# Extract legend
legend_plot <- get_legend_plot(dcifer_plot)

# View legend
grid.newpage()
grid.draw(legend_plot)

# Save legend separately

ggsave("Final_Manuscript_Figures/Re-analysis_Figs_2026/Kamwezi_dcifer_legend.png",
       plot = arrangeGrob(legend_plot),
       width = 3.5, height = 6, dpi = 300, bg = "white")
```


#### Color by Resistance Markers
```{r,}
### P44IL
tg_with_coords <- tg_with_coords %>%
  activate(nodes) %>%
  left_join(k13_P441L, by = c("name" = "id"))

tg_with_coords %>% 
  activate(nodes) %>%
  as_tibble() %>% 
  select(name, k13_P441L) %>% 
  head()

dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.5,
    color = "gray50",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = k13_P441L),
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "WT" = "#1b9e77",
      "Mix or Mutant" = "#d95f02"
    ),
    na.value = "gray80",
    name = "k13_P441L"
  ) +
  labs(title = "Kamwezi k13_P441L Mutation", color = "Genotype") +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

dcifer_plot

ggsave("DCIFER_RESULTS/Kamwezi_dcifer_Clusters_fixed_P441L.png", dcifer_plot, width = 10, height = 8, units = "in")
#--------------------------------------------------------------------------
## R561H
tg_with_coords <- tg_with_coords %>%
  activate(nodes) %>%
  left_join(k13_R561H, by = c("name" = "id"))

tg_with_coords %>% 
  activate(nodes) %>%
  as_tibble() %>% 
  select(name, k13_R561H) %>% 
  head()

dcifer_plot <- ggraph(tg_with_coords, layout = "manual", x = x, y = y) +
  geom_segment(
    data = short_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    alpha = 0.5,
    color = "gray50",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_node_point(
    aes(color = k13_R561H),
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "WT" = "#1b9e77",
      "Mix or Mutant" = "#d95f02"
    ),
    na.value = "gray80",
    name = "k13_R561H"
  ) +
  labs(title = "Kamwezi k13_R561H Mutation", color = "Genotype") +
  coord_fixed() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

dcifer_plot

ggsave("DCIFER_RESULTS/Kamwezi_dcifer_Clusters_fixed_R561H.png", dcifer_plot, width = 10, height = 8, units = "in")
##------------------------------------------------------------------------------
```



